---
title: "Lab4"
author: "Denys Kushnirenko"
date: "2023-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## 1. For three (or more) gradations of a qualitative variable, output the values for each gradation in the form of line charts (or boxplots) and, if possible, the mean values (preferably in different colors and with labels)

```{r, echo=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)
library(qqplotr)
library(ggpubr)
library(nortest)
dataset <- read.csv(".\\fatalities_isr_pse_conflict_2000_to_2023.csv")
dataset <- dataset %>%   
  filter(citizenship == 'Israeli' | citizenship == 'Palestinian' | citizenship == 'Jordanian')
dataset$citizenship <- factor(dataset$citizenship)
```
```{r}
boxplot(age ~ citizenship, data = dataset,
        xlab = "Citizenship", ylab = "Age", col = c("red", "green", "blue"))

```

## 2. Check the fulfillment of the assumptions of the ANOVA.

Check the normality of the model residuals:

```{r}
mod1 <- aov(age ~ citizenship, data = dataset)
residuals <- mod1$residuals
hist(mod1$residuals, main = "Histogram of the model residuals", freq = F, col = "green1")
ggqqplot(mod1$residuals, color = "green3")
length(mod1$residuals)
lillie.test(mod1$residuals)
```
We see a noticeable deviation from the normal distribution of the model residuals. We can conduct a rank-based analysis of variance by Kruskal-Wallis - a counterpart of the classical one-way analysis of variance.
Check the homogeneity of variances:
```{r}
tapply(dataset$age, dataset$citizenship, var, na.rm=T)
```
The condition of homogeneity of variances is also not met.

## 3. Conduct an ANOVA test for that data

ANOVA by Kruskal-Wallis:

```{r}
kruskal.test(age ~ citizenship, data = dataset)

```
Conduct an ANOVA test with Welch's correction:

```{r}
oneway.test(age ~ citizenship, data = dataset)
```
Based on the results of these studies, we have grounds to reject the null hypothesis of the equality of mathematical expectations in all groups of observations.
